|

### 2023-07-07 src/manuscript.qmd
send off a draft of this

Are White identifiers those who are percieve economic threats or is it a backlash to a loss of influence over politics? The existing work offers an unclear answer to this question. This note attempts to provide an ombnibus test to determine not only which of these questions has more empirical support but also which among common proxies of economic and political power in the United States appear to be reliable predictors of a White individual's expression of racial identity.

A common source of data and of measures for researchers in the United States is the American National Election Study. I use a penalized regression using data from the 2012, 2016, and 2020 American National Election Studies to examine the degree to which individual measures of economic and political influence explain an individual's uptake of White identity as well as to examine the degree to which changes in the political context may increase or decrease the explanatory power of these measures. I find that proxies reporting concerns about Whites' political influence are stronger explantory factors than proxies of economic influence in the post-Trump era.

@@

Social identity theory, and its offshoots, would suggest an alternative perspective: that Whites expressing a racial identity are responding to political, cultural, and societal threats. While these threats are a bit more hard to define, there are a number of examples of Whites expressing prejudice towards other racial groups in response to these sorts of challenges to their power. @tesler_2016 demonstrates that Whites heavily racialized American politics in response to Barack Obama's ascention to the White House. Other work demonstrates that the saliency and reprocussions of aligning with competing racial groups drove non-White political candidates to distance themselves from their racial confederates in an attempt to appease Whites [@stephens-dougan_2020]. Some pundits on Fox News suggest that Whites are in direct competition for jobs with immigrants from South and Central America. Despite these claims, some evidence suggests that Whites' hostility directed toward non-White immigrant groups depends on the degree to which they assimilate [@green_staerkle_2013; see also @hainmueller_hopkins_2014]. In terms of attitudes oriented toward the in-group, @dawson_1994_pup argues that Black Americans hold a strong racial identity that supercedes their class-based identity due to the persistent focus on Black individual's race in their personal and political life. For Whites, their racial identity has long been central to their behavior and the way in which they mobilize to shape legal and political institutions to maintain power [@rile-hayward_2013].

@@

Identity is largely an elusive concept to operationalize. Though there are implicit measures of White identity [see @knowles_peng_2005_jpsp], the common and measure of white identity is a response to the question, "How important is being White to your identity?" This measure captures a significant feature of an identity: the centrality or the importance of that identity to an individual. Though a rather rough measure of something as complex as identity, it has been shown to produce similar results to more complex measures of white identity and holds a significant amount of face validity [@jardina_2019_cup].

@@

The primary contribution of this paper is to examine which of a number of competing measures of different types of threat may predict white identity for individuals. As this requires an ombnibus test, I elect to use a penalized regression. I provide a more detailed discussion of the reasons why I use an estimator with an L2 regularization parameter, often refered to as a LASSO model in the supplementary materials. For a brief comment as to why I use this approach is because approach is designed to take models with a large number of predictors and to shrink the coefficients toward zero. In practice, this often means that predictors that has relatively fewer explanatory power have much smaller coefficients than the variables that do provide significant predictive power.

@@

```{r}
#| label: tbl-2012-lasso-simple
#| tbl-cap: Predictors of white Identity in 2012
model_2012 <- list(
  "2012 LWD" = lasso[["2012_simple"]]
  , "2012" = lasso[["2012_ordinal"]]
)
rows <- tribble(~term, ~`2012 LWD`, ~`2012`, 'Economic', '', '', 'Non-material', '', '', 'Demographics', '', '', 'Thresholds', '', '')
attr(rows, 'position') <- c(1, 18, 27, 32)

@@

```{r}
#| label: tbl-2012-gender-lasso
#| tbl-cap: Predictors of Whit Identity in 2012 conditional on gender

@@

```{r}
#| label: tbl-2012-pid-lasso
#| tbl-cap: Predictors of Whit Identity in 2012 conditional on partisanship

@@

```{r}
#| label: tbl-2016-lasso-simple
#| tbl-cap: Predictors of white identity in 2016
model_2016 <- list(
  "2016 LWD" = lasso[["2016_simple"]]
  , "2016" = lasso[["2016_ordinal"]]
)
rows <- tribble(~term, ~`2016 LWD`, ~`2016`, 'Economic', '', '', 'Non-material', '', '', 'Demographics', '', '', 'Thresholds', '', '')
attr(rows, 'position') <- c(1, 18, 27, 32)

@@

```{r}
#| label: tbl-2016-gender-lasso
#| tbl-cap: Predictors of Whit Identity in 2016 conditional on gender

@@

```{r}
#| label: tbl-2016-pid-lasso
#| tbl-cap: Predictors of Whit Identity in 2016 conditional on partisanship

@@

modelsummary(model_2016_pid, coef_map = cm_pid, statistic = "conf.int", gof_map = NA, add_rows = rows, notes = c('Median estimate from fitted model with 6 chains and 2000 iterations.', '95 percent credible intervals in brackets.', 'Data source: 2016 American National Election Study.'))
```
```{r}
#| label: tbl-2020-lasso-simple
#| tbl-cap: Predictors of white identity in 2020
model_2020 <- list(
  "2020 LWD" = lasso[["2020_simple"]]
  , "2020" = lasso[["2020_ordinal"]]
)
rows <- tribble(~term, ~`2020 LWD`, ~`2020`, 'Economic', '', '', 'Non-material', '', '', 'Demographics', '', '', 'Thresholds', '', '')
attr(rows, 'position') <- c(1, 18, 25, 30)

@@

```{r}
#| label: tbl-2020-gender-lasso
#| tbl-cap: Predictors of Whit Identity in 2020 conditional on gender

@@

```{r}
#| label: tbl-2020-pid-lasso
#| tbl-cap: Predictors of Whit Identity in 2012 conditional on partisanship

>>>>

Do economic or political threats motivate the expression of a racial identity among Whites in politics? The existing work offers an unclear answer to this question. This note attempts to provide an ombnibus test to determine not only which of these questions has more empirical support but also which among common proxies of economic and political power in the United States appear to be reliable predictors of a White individual's expression of racial identity.

As the American National Election Study is a long-running survey of Americans', it provides some of the most consistent data on common political attitudes and behaviors. Using data from the 2012, 2016, and 2020 administrations of the ANES, I fit an ombnibus model with penalized regression to examine the degree to which individual measures of economic and political influence explain an individual's uptake of White identity as well as to examine the degree to which changes in the political context may increase or decrease the explanatory power of these measures. From this test, I find that proxies reporting concerns about Whites' political influence are stronger explantory factors than proxies of economic influence in the post-Trump era.

@@

Social identity theory, and its offshoots, would suggest an alternative perspective: that Whites expressing a racial identity are responding to political, cultural, and societal threats. While these threats are a bit more hard to define, there are a number of examples that demonstrate Whites are responsive to cultural or political changes that present as a threat. For example, @mutz_status_2018 finds little support for the claim that the 2016 Presidential Election was a response to the economic effects of a globalizing economy and instead argues that it was a response to concerns about the changes brought about by globalization and the racial-diversity it promotes. Others demonstrate that the behavioral manifestations of a White constituency in the United States was a reflection of the saliency of race in politics [@agadjanian_lacy_2021].
A final example demonstrated that exposure to a story about discrimination about a White person lead individuals concerned with the racial hierarchy to report that Whites faced similar amounts of institutional racism as Blacks [@marshburn_et-al_2022_gpir]. Others have made similar arguments through historical lenses by arguing that while race and class are highly correlated, the saliency of race in society and in our institutions continue to make it a salient group by which we structure our thinking about who has status and power [@helms_carter_1990;@dawson_1994_pup;@rile-hayward_2013]

@@

Identity is largely an elusive concept to operationalize [see @leach_et-al_2008]. Though there are implicit measures of White identity [see @knowles_peng_2005_jpsp], the common and measure of white identity is a response to the question, "How important is being White to your identity?" This measure captures a significant feature of an identity: the centrality or the importance of that identity to an individual [@jardina_2019_cup]. Though a rather rough measure of something as complex as identity, it has been shown to produce similar results to more complex measures of white identity and holds a significant amount of face validity [@jardina_2019_cup].

@@

The primary contribution of this paper is to examine which, of several competing measures, of different types of threat may predict white identity for individuals. As this requires an ombnibus test, I elect to use a penalized regression. I provide a more detailed discussion of the reasons why I use an estimator with an L2 regularization parameter, often refered to as a LASSO model in the supplementary materials. In practice, the model applies a penalization term so predictors that have relatively fewer explanatory power have much smaller coefficients than the variables that do provide significant predictive power.

@@

```{r}
#| label: tbl-2012-lasso-simple
#| tbl-cap: Predictors of White Identity in 2012
model_2012 <- list(
  "2012 LWD" = lasso[["2012_simple"]]
  , "2012" = lasso[["2012_ordinal"]]
)
rows <- tribble(~term, ~`2012 LWD`, ~`2012`, 'Economic', '', '', 'Non-material', '', '', 'Demographics', '', '', 'Thresholds', '', '')
attr(rows, 'position') <- c(1, 18, 27, 32)

@@

```{r}
#| label: tbl-2012-gender-lasso
#| tbl-cap: Predictors of White Identity in 2012 conditional on gender

@@

```{r}
#| label: tbl-2012-pid-lasso
#| tbl-cap: Predictors of White Identity in 2012 conditional on partisanship

@@

```{r}
#| label: tbl-2016-lasso-simple
#| tbl-cap: Predictors of White Identity in 2016
model_2016 <- list(
  "2016 LWD" = lasso[["2016_simple"]]
  , "2016" = lasso[["2016_ordinal"]]
)
rows <- tribble(~term, ~`2016 LWD`, ~`2016`, 'Economic', '', '', 'Non-material', '', '', 'Demographics', '', '', 'Thresholds', '', '')
attr(rows, 'position') <- c(1, 18, 27, 32)

@@

```{r}
#| label: tbl-2016-gender-lasso
#| tbl-cap: Predictors of White Identity in 2016 conditional on gender

@@

```{r}
#| label: tbl-2016-pid-lasso
#| tbl-cap: Predictors of White Identity in 2016 conditional on partisanship

@@

modelsummary(model_2016_pid, coef_map = cm_pid, statistic = "conf.int", gof_map = NA, add_rows = rows, notes = c('Median estimate from fitted model with 6 chains and 2000 iterations.', '95 percent credible intervals in brackets.', 'Data source: 2016 American National Election Study.'))
```
```{r}
#| label: tbl-2020-lasso-simple
#| tbl-cap: Predictors of White identity in 2020
model_2020 <- list(
  "2020 LWD" = lasso[["2020_simple"]]
  , "2020" = lasso[["2020_ordinal"]]
)
rows <- tribble(~term, ~`2020 LWD`, ~`2020`, 'Economic', '', '', 'Non-material', '', '', 'Demographics', '', '', 'Thresholds', '', '')
attr(rows, 'position') <- c(1, 18, 25, 30)

@@

```{r}
#| label: tbl-2020-gender-lasso
#| tbl-cap: Predictors of White Identity in 2020 conditional on gender

@@

```{r}
#| label: tbl-2020-pid-lasso
#| tbl-cap: Predictors of White Identity in 2012 conditional on partisanship
